---
title: "HW1: A Quarto tutorial + intro to R"
author: "Hangyu Li"
format: pdf
---

# Practice Exercises with Palmer Penguins

## Setup (5 points)

Load the required package and examine the data:

```{r}
library(palmerpenguins)
```

```{r}
# Your code here: Explore the dataset structure
# How many rows? How many columns? What variable types?

penguins <- palmerpenguins::penguins
print(paste("It has", dim(penguins)[1], "rows."))
print(paste("It has", dim(penguins)[2], "colomns."))
str(penguins)
```


## Part 1: Vectors and Summary Statistics (25 points)

### 1.1 Extract and analyze bill lengths

```{r}
# Extract bill_length_mm as a vector
#bill_lengths <- # Your code here
bill_lengths <- penguins$bill_length_mm

# Calculate: mean, median, sd, min, max (handle NAs!)
print(paste("The mean is", mean(bill_lengths, na.rm = TRUE)))
print(paste("The median is", median(bill_lengths, na.rm = TRUE)))
print(paste("The sd is", sd(bill_lengths, na.rm = TRUE)))
print(paste("The min is", min(bill_lengths, na.rm = TRUE)))
print(paste("The max is", max(bill_lengths, na.rm = TRUE)))
```

### 1.2 Logical subsetting

```{r}
# Create logical vector: TRUE when bill length > 45mm
#long_bills <- # Your code here
long_bills <- bill_lengths > 45

# How many penguins have long bills?
print(paste(sum(long_bills, na.rm = TRUE), "penguins have long bills"))

```



## Part 2: Subsetting Data (25 points)

### 2.1 Island analysis

```{r}
# Extract bill lengths for Dream island penguins only
# Calculate the mean for this subset
dream_bill_length <- penguins[penguins$island == "Dream",]$bill_length_mm
print(paste("The mean of the subset is", mean(dream_bill_length, na.rm = TRUE)))
```

### 2.2 Species counting

```{r}
# How many Adelie penguins are in the dataset?
print(paste(nrow(penguins[penguins$species == "Adelie",]),
            "Adelie penguins are in the dataset"))
```

### 2.3 Find the heaviest penguin

```{r}
# Find: body mass, species, and island of the heaviest penguin
penguins[penguins$body_mass_g == max(penguins$body_mass_g, na.rm = TRUE) & 
  !is.na(penguins$body_mass_g), 
  c("body_mass_g", "species", "island")]
```



## Part 3: Matrices and apply() (25 points)

### 3.1 Create measurement matrix

```{r}
# Remove rows with missing values
#penguins_complete <- penguins[complete.cases(penguins), ]
penguins_complete <- penguins[complete.cases(penguins), ]

# Create matrix with: bill_length_mm, bill_depth_mm, 
# flipper_length_mm, body_mass_g
#penguin_measurements <- # Your code here
penguin_measurements <- as.matrix(penguins_complete[,c("bill_length_mm", 
  "bill_depth_mm", "flipper_length_mm", "body_mass_g")])
```

### 3.2 Apply functions

```{r}
# Use apply() to calculate mean of each column
apply(penguin_measurements, 2, mean)

# Use apply() to find which row has the maximum for each column
apply(penguin_measurements, 2, which.max)
```



## Part 4: Custom Function (20 points)

### 4.1 Write standardize function

```{r}
# Function that: removes NAs, returns (x - mean) / sd
#standardize <- function(x) {
  # Your code here
#}
standardize <- function(x) {
  x_complete <- x[!is.na(x)]
  return ((x_complete - mean(x_complete))/sd(x_complete))
}


# Test on bill_length_mm
#bill_length_standardized <- standardize#(penguins$bill_length_mm)
bill_length_standardized <- standardize(penguins$bill_length_mm)

# Verify: mean ≈ 0, sd ≈ 1
print(paste("The standardized mean is", mean(bill_length_standardized)))
print(paste("The standardized sd is", sd(bill_length_standardized)))
```



## Bonus (10 points max)

Choose ONE bonus challenge:

**Option A:** Create a summary data frame with one row per species showing: count, mean body mass, mean flipper length

**Option B:** Calculate bill length-to-depth ratio for each penguin. Which species has the highest average ratio?

```{r}
# Your bonus code here
library(tidyverse)

# Option A
penguins |> 
  select(species, body_mass_g, flipper_length_mm) |> 
  group_by(species) |> 
  summarise(count = n(),
            mean_body_mass = mean(body_mass_g, na.rm = TRUE),
            mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE)
  )

# Option B
penguins <- penguins |> 
  mutate(l2d = bill_length_mm / bill_depth_mm)


penguins |> 
  select(species, l2d) |> 
  group_by(species) |> 
  summarise(mean_l2d = mean(l2d, na.rm = TRUE)) |> 
  arrange(desc(mean_l2d))
```

Gentoo has the highest average bill length-to-depth ratio.


