---
title: "HW2: Data Import, Cleaning, and Exploration"
author: "Hangyu Li"
date: today
format:
  html:
    toc: true
    embed-resources: true
---

## Part 0: Set Up Your Project in Positron (5 points)

### 0.1 Create a project folder

1.  In Positron, choose **File → Open Folder…** and create/open a new folder called `STAT204-HW2` (you can put it inside the Git repository you set up in Week 1).
2.  In the **Explorer** pane (left sidebar), create a subfolder called `data`.

Opening the folder matters: Positron’s console and Quarto both use it as the **working directory**, so relative paths like `data/...` just work. **Do not use `setwd()`** in your document.

### 0.2 Download the dataset

1.  Go to: https://datadryad.org/stash/dataset/doi:10.5061/dryad.ts45225
2.  Click “Download Dataset” (free and open access) and extract the files.
3.  Find `PalmTraits_1.0.txt` (a tab-delimited text file, under the Sep 09, 2020 tab) and move it into your `data/` folder.

### 0.3 Create your Quarto document

1.  Choose **File → New File… → Quarto Document**, and save it in the project folder (not in `data/`) as `LastName_FirstName_HW2.qmd`.
2.  Check the **interpreter picker** in the top-right corner of Positron: it should show an **R** session. If it shows Python, click it and select R.
3.  Replace the default header with:

```{yaml}
---
title: "HW2: Data Import, Cleaning, and Exploration"
author: "Your Name"
date: today
format:
  html:
    toc: true
    embed-resources: true
---
```

4.  Add a short paragraph saying where the data comes from (the Dryad link and the file name), then this setup chunk:

```{r}
library(dplyr)   # used in Part 3.3 only
```

**Working in Positron:**

- Run the current line or selection with **Ctrl/Cmd + Enter**, or a whole chunk with **Run Cell** above it.
- The **Variables** pane shows every object you create. Click the table icon next to a data frame (or run `View(palm_data)` in the console) to open it in the **Data Explorer**.
- **Preview** your document often (the Preview button, or **Ctrl/Cmd + Shift + K**). If it renders, your code runs from top to bottom in a fresh session, which is what “reproducible” means.
- Optional: commit your work to Git from the **Source Control** pane as you go, like you practiced in Week 1.

**Points:** the document renders without errors, uses a relative path for the data, and has a setup chunk.

------------------------------------------------------------------------

## Part 1: Import and First Look (15 points)

### 1.1 Import the data (8 points)

```{r}
# palm_data <- read.delim("data/PalmTraits_1.0.txt",
#                         header = TRUE,
#                         stringsAsFactors = FALSE)
```

**Instructions:**

- Import the file with `read.delim()` (it is tab-delimited, which is the default for `read.delim()`)
- Show `dim(palm_data)`, `names(palm_data)`, and `head(palm_data, 5)`
- Show `str(palm_data)`
- In one or two sentences: how many species and variables are there? Which variables did R read as numbers, and which as text?

### 1.2 Guided exploration in the Data Explorer (7 points)

Open `palm_data` in Positron’s **Data Explorer** (Variables pane → table icon). Use the column headers to sort and the filter bar to filter. The summary panel on the left shows the percentage of missing values for each column.

Answer these in your document, in text:

1.  Sort by `MaxStemHeight_m` from largest to smallest. Which species is the tallest, and what is its growth form (look at `Climbing`)? Does a “height” of that size make sense for that growth form?

2.  Filter to `Climbing` equal to 1. How many climbing species are there?

3.  Which of these columns has the most missing values: `MaxStemHeight_m`, `Max_Blade_Length_m`, `AverageFruitLength_cm`, `AverageFruitWidth_cm`?

4.  Look at the `UnderstoreyCanopy` column. Besides `understorey` and `canopy`, what other values appear? Is anything missing that R would **not** count as `NA`?

Then confirm (a) and (b) with code, so the answers appear in your rendered document:

```{r}
# palm_data$SpecName[which.max(palm_data$MaxStemHeight_m)]
# sum(palm_data$Climbing == 1)
# table(palm_data$UnderstoreyCanopy)
```

------------------------------------------------------------------------

## Part 2: Data Cleaning with Base R (35 points)

### 2.1 Find the missing values (8 points)

```{r}
# Missing values in the whole dataset, and per column
# sum(is.na(palm_data))
# colSums(is.na(palm_data))

# Percentage missing per column, sorted
# pct_missing <- round(100 * colMeans(is.na(palm_data)), 1)
# sort(pct_missing, decreasing = TRUE)

# Duplicated rows
# sum(duplicated(palm_data))
```

**Instructions:** run the code above, then write two or three sentences: which kinds of traits (stem, leaf, fruit) have the most missing data, and are there duplicated rows?

### 2.2 Create a working subset (5 points)

Keep only these columns:

```{r}
# keep <- c("SpecName", "Climbing", "Acaulescent", "Erect", "StemArmed",
#           "MaxStemHeight_m", "Max_Blade_Length_m",
#           "AverageFruitLength_cm", "AverageFruitWidth_cm")
# palm_subset <- palm_data[, keep]
# dim(palm_subset)
```

Report how many variables and observations your subset has.

### 2.3 Handle missing values (8 points)

We will keep the species that have **both** a stem height and a fruit length, since Parts 3 and 4 use those two variables the most:

```{r}
# key_vars <- c("MaxStemHeight_m", "AverageFruitLength_cm")
# palm_clean <- palm_subset[complete.cases(palm_subset[, key_vars]), ]
# nrow(palm_subset) - nrow(palm_clean)   # how many rows were removed?
# colSums(is.na(palm_clean))             # what is still missing?
```

**Instructions:** run the code and answer in two or three sentences:

- How many species did you remove, and what percentage of the data is that?
- Which variables still have missing values? (That’s fine: use `na.rm = TRUE` when you summarize them.)
- Could removing these species bias the results? Think about which kinds of palms are more likely to be missing measurements.

### 2.4 Fix data types (6 points)

`StemArmed` is stored as a number, but it is really a category. Convert it to a factor with readable labels, and check the result:

```{r}
# str(palm_clean$StemArmed)
# palm_clean$StemArmed <- factor(palm_clean$StemArmed,
#                                levels = c(0, 1),
#                                labels = c("Unarmed", "Armed"))
# table(palm_clean$StemArmed, useNA = "ifany")
```

**Instructions:** do the same for `Climbing`, using `levels = c(0, 1, 2)` and `labels = c("No", "Yes", "Both")`. Show `str(palm_clean)` after your conversions, and explain in one or two sentences why it matters that these variables are factors rather than numbers (think about what `mean()` or `summary()` would give you).

### 2.5 Create new variables (8 points)

1.  **Size category** from stem height. Run this code and show the table:

```{r}
# palm_clean$size_category <- ifelse(palm_clean$MaxStemHeight_m <= 5, "Small",
#                             ifelse(palm_clean$MaxStemHeight_m <= 15, "Medium",
#                             ifelse(palm_clean$MaxStemHeight_m <= 30, "Tall", "Very Tall")))
# palm_clean$size_category <- factor(palm_clean$size_category,
#                                    levels = c("Small", "Medium", "Tall", "Very Tall"))
# table(palm_clean$size_category)
```

2.  **Fruit aspect ratio**: create `fruit_ratio` = `AverageFruitLength_cm` / `AverageFruitWidth_cm`. Use `summary()` on it. What does a ratio above 1 mean about the shape of the fruit? How many values are `NA`, and why?

------------------------------------------------------------------------

## Part 3: Descriptive Statistics (20 points)

### 3.1 Summary statistics (10 points)

For `MaxStemHeight_m` **and** `AverageFruitLength_cm`, calculate the mean, median, standard deviation, and the quartiles:

```{r}
# x <- palm_clean$MaxStemHeight_m
# mean(x); median(x); sd(x)
# quantile(x, probs = c(0, 0.25, 0.5, 0.75, 1))
```

Organize the results in a small data frame with one row per variable and one column per statistic, and print it.

**Interpret:** For each variable, compare the mean and the median. What does the difference tell you about the shape of the distribution?

### 3.2 Frequency tables (5 points)

```{r}
# table(palm_clean$size_category)
# prop.table(table(palm_clean$size_category))

# Two-way table: size category vs armature, with row proportions
# tab <- table(palm_clean$size_category, palm_clean$StemArmed)
# tab
# round(prop.table(tab, margin = 1), 2)
```

**Interpret:** Which size category has the highest proportion of armed (spiny) palms? Write one sentence.

### 3.3 Group summaries with `dplyr` (5 points)

`group_by()` splits the data into groups, and `summarize()` calculates statistics for each group. The pipe `%>%` (read it as “then”) passes the result of one step to the next:

```{r}
# palm_clean %>%                       # start with the clean data, THEN
#   group_by(StemArmed) %>%            # group it by armature, THEN
#   summarize(
#     n = n(),                                             # species per group
#     mean_height = mean(MaxStemHeight_m),
#     sd_height = sd(MaxStemHeight_m),
#     median_fruit = median(AverageFruitLength_cm)
#   )
```

**Instructions:** run this summary, then change it to group by `size_category` and add **one more statistic** of your choice (for example `max()` of stem height, or the median of `Max_Blade_Length_m` with `na.rm = TRUE`). Store the result as `palm_summary` and print it. (If you see an `NA` group, those are species whose armature is unknown.) Write one or two sentences about what it shows.

------------------------------------------------------------------------

## Part 4: Visualizations with Base R (15 points)

For each plot, add a title and axis labels with units, and write **one or two sentences** interpreting it.

### 4.1 Histogram of stem height (5 points)

Use `hist()` on `MaxStemHeight_m`. Describe the shape. Then make a second histogram of `log10(MaxStemHeight_m + 1)` and compare. (Why `+ 1`? Look for zeros in the data.)

### 4.2 Boxplot of stem height by armature (5 points)

Use `boxplot(MaxStemHeight_m ~ StemArmed, data = palm_clean, ...)`. Are armed palms taller or shorter than unarmed palms, typically? Are there many outliers?

### 4.3 Scatterplot of fruit dimensions (5 points)

Plot `AverageFruitWidth_cm` (y) against `AverageFruitLength_cm` (x). Add the line `abline(0, 1)`, where length equals width. What does it mean for a species to be above or below that line? Use `cor(..., use = "complete.obs")` to report the correlation.

------------------------------------------------------------------------

## Part 5: Reflection (10 points)

Write about **150 words** addressing:

1.  **Cleaning decisions:** What was the biggest data quality issue you found, and how did your choices in Part 2 change the data you analyzed?
2.  **Findings:** What are the two most interesting patterns you found about palm size, armature, or fruit shape?
3.  **Workflow:** What was one thing that Positron’s Data Explorer or rendering your Quarto document helped you catch?

------------------------------------------------------------------------

## Optional Extensions

These are **not graded**. They are good practice for HW3 and the midterm. If you do them, put them in a section called “Optional” at the end of your document.

1.  **Outliers.** For `MaxStemHeight_m`, compute Q1, Q3, and the IQR with `quantile()`, and flag values above Q3 + 1.5 × IQR. List the flagged species. Are they errors, or real extreme values? (Some palms, like *Ceroxylon quindiuense*, exceed 60 m, and climbing *Calamus* can exceed 100 m in length.) Decide whether to keep, remove, or flag them, and explain why.
2.  **A primary growth form variable.** Combine `Climbing`, `Acaulescent`, and `Erect` into a single factor with levels “Climbing”, “Acaulescent”, “Erect”, and “Mixed”, using nested `ifelse()`. Make a frequency table, then redo the boxplot in 4.2 grouped by growth form.
3.  **More pipes.** Use `%>%` to (a) keep only climbing palms, (b) select three columns and show the first 5 rows, and (c) compute the mean stem height.
4.  **Grouping by two variables.** Repeat 3.3 with `group_by(size_category, StemArmed)`.
5.  **Base R vs. dplyr.** Reproduce your `palm_summary` means with `tapply()` or `aggregate()`. Which approach do you find easier to read?
6.  **Multi-panel figure.** Use `par(mfrow = c(2, 2))` to combine your plots and a `barplot()` of size-category frequencies into one figure.

------------------------------------------------------------------------
